Model Order Selection and Eigen Similarity based Framework for Detection and Identification of Network Attacks

### 2016-11-02 docs/papers/2016JNCA/2016JNCA.tex
version with results discussion

This paper is organized as follows. In Section \ref{sec:relatedworks}, related works are discussed. Section \ref{sec:datamodel} presents the data model. Section \ref{sec:prop_getv} describes the proposed framework for blind and automatic detection of malicious traffic. Section \ref{sec:experimentalresults} discusses the experimental validation and presents the experimental results. Section \ref{sec:conclusionandfutureworks} draws the conclusions and the suggestions for future work. The \ref{sec:mos} presents mathematical concepts of examples of state-of-the-art MOS schemes.

@@

Signal processing techniques have been successfully applied to network anomaly detection \cite{Lu2009}. Lu and Ghorbani \cite{Lu2009} proposed a network anomaly detection model based on network flow, wavelet approximation, and system identification theory. However, their work did not address problems without significant outliers, such as port scan attacks. Zonglin \emph{et al} \cite{Zonglin2009} proposed a method to detect traffic anomaly with correlation analysis, where the correlation between traffic signals and the predicted traffic signals are used to reveal anomalies. Zonglin \emph{et al.} \cite{Zonglin2009} evaluated the correlation analysis for anomaly detection, but the work was not applied to probe and flood attack detection, simultaneously.

@@

This section presents the performed experiments and the acquired results. First, in Section \ref{sec:AnalyzedScenario}, the scenario adopted in the experiments is summarized. Then, Section \ref{sec:largesteigenvaluesanalysis} shows the results of the largest eigenvalue analysis by time frames. In Section \ref{sec:MOSSchemesEvaluation} are described the results of the evaluated MOS schemes for attack detection. Section \ref{sec:EigenvalueAnalysis} presents the results of the eigenvalue analysis for identification of time frames under attack, Section \ref{sec:EigenSimilarityAnalysis} shows the results of similarity analysis for detailed flood and port scan identification, and Section \ref{sec:Complexity} discuss the computational complexity of the proposed framework.

\subsection{Analyzed Scenario}
\label{sec:AnalyzedScenario}

@@

\subsection{Complexity Analysis}
\label{sec:Complexity}

This subsection discusses the computational complexity of the proposed framework, focusing on the main steps, which are the eigenvalues decomposition (EVD), largest eigenvalues analysis, application of MOS scheme and eigen similarity analysis, according to Figure \ref{fig:fig80} and equations presented in Section \ref{sec:prop_getv}.

@@

>>>>

This paper is organized as follows. In Section \ref{sec:relatedworks}, related works are discussed. Section \ref{sec:datamodel} presents the data model and the evaluated datasets. Section \ref{sec:prop_getv} describes the proposed framework for blind and automatic detection of flood and probe attacks. Section \ref{sec:experimentalresults} discusses the experimental validation and presents the results, and Section \ref{sec:Complexity} discuss the computational complexity of the proposed framework and evaluates the required processing time for simulated scenarios. Section \ref{sec:conclusionandfutureworks} draws the conclusions and the suggestions for future work. The \ref{sec:mos} presents mathematical concepts of examples of state-of-the-art MOS schemes.

@@

Signal processing techniques have been successfully applied to network anomaly detection \cite{Lu2009}. Lu and Ghorbani \cite{Lu2009} proposed a network anomaly detection model based on network flow, wavelet approximation, and system identification theory. However, their work requires a training method to produce a prediction model for normal daily traffic and presents limitations on identification of behaviors without significant outliers, such as port scan attacks. Zonglin \emph{et al} \cite{Zonglin2009} proposed a method to detect traffic anomaly with correlation analysis, where the correlation between traffic signals and the predicted traffic signals are used to reveal anomalies. Zonglin \emph{et al.} \cite{Zonglin2009} evaluated the correlation analysis for anomaly detection, but the work was not applied to probe and flood attack detection, simultaneously.

@@

This section presents the performed experiments and the acquired results. First, in Section \ref{sec:AnalyzedScenario}, the simulated scenario adopted in the experiments is summarized. Then, Section \ref{sec:largesteigenvaluesanalysis} shows the results of the largest eigenvalue analysis by time frames for the simulated scenario. In Section \ref{sec:MOSSchemesEvaluation} are described the results of the evaluated MOS schemes for attack detection in the simulated dataset. Section \ref{sec:EigenvalueAnalysis} presents the results of the eigenvalue analysis for identification of time frames under attack, Section \ref{sec:EigenSimilarityAnalysis} shows the results of similarity analysis for detailed flood and port scan identification for the simulated scenario. Section \ref{sec:DarpaEvaluation} presents the results of the largest eigenvalue analysis, model order selection and the eigenvalue analysis for flood and probe attack detection in the DARPA 1998 dataset.

\subsection{Simulated Scenario}
\label{sec:AnalyzedScenario}

@@

\subsection{DARPA Scenario}
\label{sec:DarpaEvaluation}

This subsection presents a summarized view of results obtained from the application of the proposed framework, focusing on the largest eigenvalue analysis, model order selection and the eigenvalue analysis, for flood and probe attack detection in the DARPA 1998 dataset. Since the proposed framework is detailed explained in previous sections, here we focus on the parameter selection, dataset evaluation and results for flood and probe attack identification.

The DARPA dataset includes 7 weeks of sniffed traffic saved into raw packet data. The traffic and the labeled attacks are grouped by week and day, with information of the number and types of attacks per day, but also providing the start time for each labeled attack. For this evaluation, we performed an evaluation per day, considering the network traffic of 24 hours split into $Q$ time frames of 60 minutes ($N = 60$) and aggregated by minute and by port number. For each time frame $q$, a traffic matrix $\boldsymbol{X}^{(q)} \in \mathbb{R}^{17 \times 20}$ was obtained, considering the ports 20, 21, 22, 23, 25, 79, 80, 88, 107, 109, 110, 113, 115, 143, 161, 389 and 443.

Since the proposed framework focus on flood and probe attack detection, we evaluated only the attacks with behavior similar to flood or probe attack. Initially we selected all DoS and probe attacks, but we observed that the most cases of DoS focus on exploit system vulnerabilities instead of flooding attack, and most of probe attacks focus on ICMP intead of port scanning. Therefore, for evaluation of the proposed approach for flood and probe attack detection, it is necessary to select cases that implements flood or port scan behaviors. We select the following week-day-attack cases:

\begin{enumerate}
	\item week3-thursday-neptune;
	\item week4-friday-portsweep;
	\item week5-thursday-neptune;
	\item week5-thursday-portsweep;
	\item week5-friday-portsweep;
	\item week6-wednesday-neptune;
	\item week6-thursday-neptune;
	\item week7-wednesday-portsweep.
\end{enumerate}

The analysis based on sample covariation of zero mean variables was evaluated for flooding behavior of netpune attacks, obtaining rates of 100\% for true positive (TP) detection and 60\% for false positive (FP) detection from 30 time frames. The results also show 50\% of misclassification rate \cite{bhuyan2014network}, which attempts to estimate the probability of disagreement between the true and predicted cases by dividing the sum of FN and FP by the total number of pairs observed. 
The result for FP and misclassification analysis is poor due to the legitimate traffic of DARPA dataset presents high number of packets per time  from one source to one target, with no variation on IP source or target port. This observation corroborates with previous evaluations of the DARPA dataset that highlight issues regarding traffic redundancy.

The analysis based on covariance of zero mean and unitary standard deviation variables was evaluated for port scan attacks, including probe attacks and DoS attacks that send few packets for several ports in order to exploit some vulnerability. The results show rates of 76.9\% for TP detection and 23\% for FP detection from 94 time frames. The observed misclassification rate for this scenario was 32.7\%. We observed that all FN cases are probe attacks with a time delay between scanning one port and start scanning the next port, what can be called as sparse probe attacks. Cases with a delay of one minute or more were not detected by the proposed approach.

The performance of detection rate of flooding attacks is compared with the method proposed by Callegari \emph{et al} \cite{Zonglin2009}. This work is a statistical method, based on PCA, without training or learning methods, even though it relies on visual analysis for principal components selection. The best detection rate of \cite{Zonglin2009} was 82\% for detection of synthetically added flood attacks, while our proposal obtains 100\% of detection rate for detection of flood attacks of DARPA dataset.

Due to the lack of statistical techniques without training or learning methods for detection of probe attacks, we compare our approach with Lu and Ghorbani's \cite{Lu2009} proposal, which is a network anomaly detection model based on signal processing techniques that uses DARPA dataset for evaluation. The results of \cite{Lu2009} show the best detection rate of 94.67\% in terms of general attack instance detection, but shows a case case with 50\% of attack instance detection for the portsweep attack. The proposed approach presents 76.9\% of detection rate measured specifically for probe attacks, without the requirement of learning or training methods, in contrast to Lu and Ghorbani's \cite{Lu2009} work.

\section{Complexity Analysis}
\label{sec:Complexity}

This section discusses the computational complexity of the proposed framework, focusing on the main steps, which are the eigenvalues decomposition (EVD), largest eigenvalues analysis, application of MOS scheme and eigen similarity analysis, according to Figure \ref{fig:fig80} and equations presented in Section \ref{sec:prop_getv}.

@@

falar sobre as melhorias para reduzir falso positivo e também para a detecção de probe atacks disperos
